---
title: "Aneuploidy association with immune features"
output:
  html_document:
    df_print: paged
---

### Context \ 
Then we ought to find out the aneuploidy association with immune effects. We will start with CD8 infiltration, which is quantified by [MCP counter](https://github.com/ebecht/MCPcounter). We will then proceed to GZMB, GNLY, PRF1 expression as provided by corresponding TCGA RNA-seq dataset. Other two factors that we need to address from TCGA RNA-seq dataset are PDL1 expression and MHCI expression, independently. Then, we will look at the TLS Status which is quantified by TLS score calculated by an equation provided by lab. Following that is the TMB status which should be provided by TCGA dataset. Lastly, we will verify our finding using HH ovarian validation, MSC single-cell RNA seq and Lung single-cell RNA seq. \ 

There are several statistical tests we could perform to address the association. We will start with the: \ 
1. Spearman correlation of cytoband and `all_lesion_conf_0.95` + correlation test `cor.test`. Remember to adjust for multiple testing by FDR to find adjusted p-values. \ 
2. Then, perform correlation and linear regression (adjust for age/ other confounding variables) or mix effect model. \ 
3. Because our data takes the form of categorical (aneuploidy status -2, -1, 0, 1, 2) vs continuous variable (MCP counter/ expression) we shall perform Kruskal–Wallis (non-parametric) test or ANOVA (normal). Choice of test depends on preliminary test of whether the distribution of continous variable is normally distributed. \ 

#### Load Libraries and environments  \ 

```{r libraries, message=FALSE, warning=FALSE}
library(dplyr)
library(magrittr)
library(caroline)
library(readr)
library(readxl)
library(stringr)
library(survival)
library(devtools)
library(MCPcounter)
library(data.table)
```

Load the past results from the clinical_association study.     
```{r environment}
rm(list = ls())
load("/rds/general/user/ph323/home/MRes_project_1/codes/1_clinical_association/clinical_association.RData")
envFiles <- ls()
envFiles <- envFiles[grepl("^result_", envFiles)]
```

#### Control Board    
Directing to the folders in HPC.     
```{r control_board}
## Control board  -------------
HOME <- "/rds/general/user/ph323/home/MRes_project_1/"
PHENOTYPE <- "/rds/general/user/ph323/home/MRes_project_1/GISTIC2_by_genePattern/clinical_docs"
RNA_SEQ <- "/rds/general/user/ph323/home/MRes_project_1/GISTIC2_by_genePattern/rna_seq"
MUTATION <- "/rds/general/user/ph323/home/MRes_project_1/docs/GISTIC2/tcga/ova/"
SCNA <- "/rds/general/user/ph323/home/MRes_project_1/GISTIC2_by_genePattern/raw_SCNA_scores"
```

### Immune Association 
#### Cleaning RNA_seq dataframe     
The dataframe is downloaded from UCSC XENA titled [gene expression RNAseq - Batch effects normalized mRNA data](https://tcga-pancan-atlas-hub.s3.us-east-1.amazonaws.com/download/EB%2B%2BAdjustPANCAN_IlluminaHiSeq_RNASeqV2.geneExp.xena.gz), with unit `log2(norm_value+1)`.    
Upon inspection, there is a duplicated row with sample called `SLC35E2`. When averaging the mean expression across samples by this code `apply(as.matrix(RNA_seq[which(RNA_seq$sample == "SLC35E2"), 2:ncol(RNA_seq)]), 1, mean)`, we found that row 16302 has value of 10.815587 and row 16303 has value of 5.472321. We enquired the ncbi gene database and found that this was likely due to the fact that SLC35E2 exist as 2 identity SLC35E2A (pseudogene) and SLC35E2B (protein coding) in human. The full name is solute carrier family 35 member E2A or E2B. On average, by looking at the expression bargraph from (1) HPA RNA-seq normal tissue, (2) RNA sequencing of total RNA from 20 human tissues, (3) Illumina bodyMap2 transcriptome, or (4) Tissue-specific circular RNA induction during human fetal developemnt. The expression level of [SLC35E2B](https://www.ncbi.nlm.nih.gov/gene/728661#gene-expression) is higher than [SLC35E2A](https://www.ncbi.nlm.nih.gov/gene/9906#gene-expression) across all human tissue samples compared to SLC35E2. Therefore to resolve the conflicted name, we will rename the lower expression row 16303 as SLC35E2A and higher expression row 16302 as SLC35E2B.     
```{r rna_seq, message=FALSE, warning=FALSE}
## Load RNA_seq 
RNA_seq <- read_table(paste0(RNA_SEQ, "/tcga_target_rna_seq"), show_col_types = FALSE) ## faster read in, in tibble
RNA_seq <- as.data.frame(RNA_seq)
#RNA_seq$sample[16302] <- "SLC35E2B" 
#RNA_seq$sample[16303] <- "SLC35E2A" 
rownames(RNA_seq) <- RNA_seq$sample 
```

#### Get immune phenotype.  
1. Getting CD8 infiltration: Install the [MCP counter](https://github.com/ebecht/MCPcounter) package from github `install_github("ebecht/MCPcounter",ref="master", subdir="Source")`.    
2. Getting the TLS score: reference provided by the paper [Tertiary lymphoid structures improve immunotherapy and survival in melanoma](https://doi.org/10.1038/s41586-019-1914-8). **Question**: how is TLS signature gene identified? Is it inherently different in different types of cancers?     
3. The TMB information was downloaded from cBiportal enquiring TCGA > by cancer type > PanCaner Atlas. And the intersecting columns are isolated by this code:      
```
for(i in TCGA){
  tmb <- read_tsv(paste0(MUTATION, "/", i,"_tmb.tsv"))
  tmb <- as.data.frame(tmb)
  tmb_name <- make.names(paste0(i, "_tmb"))
  
  assign(tmb_name, tmb)
}

intersect(intersect(colnames(lung_tmb), colnames(ova_tmb)), colnames(mel_tmb))
```
4. Specific genes: c("CD274", "GZMB", "GNLY", "PRF1", "MHCI"). [MHCI in human](https://www.ncbi.nlm.nih.gov/books/NBK27156/#_A579_) is HLA-A, HLA-B, HLA-C.   

**Functions:**     
```{r immune_pheno}
## Separate out the RNA_seq from TCGA PANCAN rna-seq dataset 
immune_pheno <- function(cancer, rna_seq, mcp_pat, tls_sig, tmb_col, gene_pat, write_each = FALSE, write_uni = TRUE){
  ## cancer = cancer type, usually given as "i" in a for (i in TCGA){} loop 
  ## rna_seq = the rna_seq dataframe 
  ## write_result = TRUE or FALSE (default to FALSE), determines if the result is write to environment. 
  ## tls_signature = c("CD79B", "CD1D", "CCR6", "LAT", "SKAP1", "CETP", "EIF1AY", "RBP5", "PTGDS", "CCL19", "CCL21", "CXCL13", "CCR7", "CXCR5", "SELL", "LAMP3")
  ## tmb_col = c("Diagnosis Age", "Sex", "Fraction Genome Altered", "Mutation Count", "TMB (nonsynonymous)") 
  ## gene_pat = "^HLA.*[ABC]$|^CD274$|^GZMB$|^GNLY$|^PRF1$"
  ## mcp_pat = c("t.cell", "cd8_t.cell", "cytotoxic_lymphocytes", "b_lineage", "nk_cells", "monocytic_lineage", "myeloid_dendritic_cells", "neutrophils", "endothelial_cells", "fibroblasts")
  
  ## Read in samples and data cleaning -----------------------
  sample_names <- get(cancer, envir = .GlobalEnv)
  rna_seq <- get(rna_seq, envir = .GlobalEnv)
  sample_names <- intersect(sample_names, colnames(rna_seq)) ## this will be our new sample_names 
  rna_seq <- rna_seq[, sample_names] ## we will only preserve cancer patients we care about, this is a DATAFRAME!!
  
  ## MCP counter ---------------------------- 
  mcp_results <- MCPcounter.estimate(
    expression = rna_seq, ## dataframe rna_seq 
    featuresType = "HUGO_symbols")
  rownames(mcp_results) <- c("t.cell", "cd8_t.cell", "cytotoxic_lymphocytes", "b_lineage", "nk_cells", 
                             "monocytic_lineage", "myeloid_dendritic_cells", "neutrophils", "endothelial_cells", 
                             "fibroblasts")
  mcp_results <- as.data.frame(t(mcp_results))
  mcp_results <- select(mcp_results, all_of(mcp_pat))
  
  ## TLS status -----------------------
  tls_signature <- tls_sig  ## from literature 
  tls_rna_seq <- rna_seq[tls_signature, ] ## isolate their rna_expression
  tls_rna_seq <- as.matrix(t(tls_rna_seq)) ## transform it into a matrix 
  tls_rna_seq <- scale(tls_rna_seq, center = F) ## scale it 
  tls_rna_seq <- as.data.frame(tls_rna_seq) ## transform it back into a dataframe 
  tls_rna_seq$TLS_mean <- apply(tls_rna_seq, 1, mean) ## calculate the TLS by mean 
  
  ## TMB calculation ----------------------
  tmb <- read_tsv(paste0(MUTATION, "/", cancer,"_tmb.tsv"), show_col_types = FALSE) ## the read in is a tibble 
  tmb <- as.data.frame(tmb)
  rownames(tmb) <- tmb$`Sample ID`
  sample_names <- intersect(sample_names, rownames(tmb)) ## there are fewer patients than the original patient list. 
  tmb <- tmb[sample_names, tmb_col]
  
  ## PDL1 exp, GZMB, GNLY, PRF1, MHC1 expression ----------------------
  genes <- which(grepl(gene_pat, rownames(RNA_seq)))
  spec_gene <- rna_seq[genes, ] 
  MHCI <- colMeans(spec_gene[rownames(spec_gene) %in% c("HLA-A", "HLA-B", "HLA-C"), ], na.rm = TRUE)
  MHCII <- colMeans(spec_gene[which(grepl("^HLA-D", rownames(spec_gene))), ], na.rm = TRUE)
  spec_gene <- rbind(spec_gene, MHCI = MHCI, MHCII = MHCII)
  spec_gene <- as.data.frame(t(spec_gene), )
  spec_gene <- spec_gene[, which(!grepl("^HLA", colnames(spec_gene)))]
  
  ## immuno_phenotype -------------------------
  ## this checks if the order of the rownames are identical because cbind does not bind by rownames 
  CHECK <- identical(rownames(mcp_results), rownames(tls_rna_seq)) && identical(rownames(mcp_results), rownames(spec_gene))
  if(CHECK){
    temp <- cbind(mcp_results, tls_rna_seq$TLS_mean, spec_gene)
    immuno_phenotype <- merge(temp, tmb, by = "row.names", all.x = TRUE)
    rownames(immuno_phenotype) <- immuno_phenotype$Row.names
    immuno_phenotype$Row.names <- NULL
    immuno_phenotype$y <- NULL
  }
  
  ## Write the unified table ------------------
  if(write_uni){
    print(paste0("This is the organised immuno_phenotype table for ", cancer, ". ", 
                 "Number of samples: ", nrow(immuno_phenotype)))
    print(immuno_phenotype[1:3, ])
    
    immuno_phenotype_name <- make.names(paste0(cancer, "_immuno_pheno"))
    assign(immuno_phenotype_name, immuno_phenotype, envir = .GlobalEnv)
    
    rna_seq_name <- make.names(paste0(cancer, "_rna_seq"))
    assign(rna_seq_name, rna_seq, envir = .GlobalEnv)
  }
  
  ## Write out each separated table ----------------
  if(write_each){
    print(paste0("This is immune phenotype for ", cancer, " cancer"))
    print(paste0("This is MCP, number of samples: ", nrow(mcp_results))); print(mcp_results[1:3, ])
    print(paste0("This is TLS, number of samples: ", nrow(tls_rna_seq))); print(tls_rna_seq[1:3, ])
    print(paste0("This is TMB, number of samples: ", nrow(tmb))); print(tmb[1:3, ])
    print(paste0("This is selected genes, number of samples: ", nrow(spec_gene))); print(spec_gene[1:3, ])
    
    rna_seq_name <- make.names(paste0(cancer, "_rna_seq"))
    assign(rna_seq_name, rna_seq, envir = .GlobalEnv)
    write.table(rna_seq, file = paste0(RNA_SEQ, "/", rna_seq_name, ".txt"), quote = FALSE, 
                row.names = TRUE, col.names = TRUE)
    
    mcp_name <- make.names(paste0(cancer, "_mcp_results"))
    assign(mcp_name, mcp_results, envir = .GlobalEnv)
    
    tls_rna_seq_name <- make.names(paste0(cancer, "_tls"))
    assign(tls_rna_seq_name, tls_rna_seq, envir = .GlobalEnv)
    
    tmb_name <- make.names(paste0(cancer, "_tmb"))
    assign(tmb_name, tmb, envir = .GlobalEnv)
    
    spec_gene_name <- make.names(paste0(cancer, "_spec_gene"))
    assign(spec_gene_name, spec_gene, envir = .GlobalEnv)
  }
}
```

```{r apply_immune_pheno}
tls_signature = c("CD79B", "CD1D", "CCR6", "LAT", "SKAP1", "CETP", "EIF1AY", "RBP5", "PTGDS", "CCL19", "CCL21", "CXCL13", "CCR7", "CXCR5", "SELL", "LAMP3")
tmb = c("Fraction Genome Altered", "TMB (nonsynonymous)") 
gene_pattern = "^HLA.*[ABC]$|^CD274$|^GZMB$|^GNLY$|^PRF1$"
tcells <- c("t.cell", "cd8_t.cell")

for(i in TCGA){
  immune_pheno(cancer = i, rna_seq = "RNA_seq", mcp_pat = tcells, tls_sig = tls_signature, 
               tmb_col = tmb, gene_pat = gene_pattern, write_each = FALSE, write_uni = TRUE)
}
```

#### Statistical test 
```{r}
statistics_test <- function(cancer, focal = FALSE, broad = FALSE, not_gistic = FALSE, test){
  ## cancer = lung, ova, or mel 
  ## test = either spearman or linear_regression
  
  ## build a list
  immuno <- list()
  
  ## get needed data 
  if(focal){cnv <- focal_cleaning(cancer)}
  if(broad){cnv <- broad_cleaning(cancer)}
  if(not_gistic){
    cnv <- get(paste0(cancer, "_scna_as"), envir = .GlobalEnv)
    cnv <- t(cnv) %>% as.data.frame()
  }
  immuno_data <- get(paste0(cancer, "_immuno_pheno"), envir = .GlobalEnv)
  pheno <- phenotype_cleaning(cancer, confounders = c("^age_", "_stage$"), col_names = c("age", "stage"))
  index <- intersect(rownames(immuno_data), colnames(cnv)) ## select out common samples 
  
  ## further process 
  pheno <- pheno[index, ]
  immuno_data <- as.matrix(immuno_data[index, ])
  cnv <- as.matrix(cnv[, index]) 
  col_names <- colnames(immuno_data)
  
  if(test == "spearman"){ ## perform spearman statistical test 
    ## run test 
    for(y in 1:ncol(immuno_data)){
      ## set up empty dataframe
      results <- array(NA, c(nrow(cnv), 2))
      rownames(results) <- rownames(cnv)
      colnames(results) <- c("cor.val", "p_values")
      results <- as.data.frame(results) ## set it as a dataframe
      
      for(x in 1:nrow(cnv)){ 
        temp <- cor.test(immuno_data[, y], as.numeric(cnv[x, ]), method = "spearman")
        results$cor.val[x] <- temp$estimate
        results$p_values[x] <- temp$p.value
      }
      
      ## multiple testing correlation by fdr? 
      results <- as.data.frame(results)
      results$FDR <- p.adjust(results$p_values, method = "fdr")
      results <- results[order(results$FDR, decreasing = FALSE), ]
      
      #print(paste0("finished analysing column ", y))
      immuno[[col_names[y]]] <- results
    }
    
    ## random test, let's use 1p 
    check_test <- cor.test(immuno_data[, 3], as.numeric(cnv[1, ]), method = "spearman")
  }
  
  if(test == "linear_regression"){ ## perform linear regression statistical test 
    ## run test 
    for(y in 1:ncol(immuno_data)){ #1:ncol(immuno_data) to replace 1:2
      ## set up empty dataframe
      results <- array(NA, c(nrow(cnv), 5))
      rownames(results) <- rownames(cnv)
      colnames(results) <- c("estimate", "std_error", "t_value", "p_values", "r_sq")
      results <- as.data.frame(results) ## set it as a dataframe
      
      for(x in 1:nrow(cnv)){ #1:nrow(cnv)
        temp <- lm(immuno_data[, y]~cnv[x, ]+pheno$age+pheno$stage, data = as.data.frame(cnv))
        temp <- summary(temp)
        results$estimate[x] <- temp$coefficients[2, 1]
        results$std_error[x] <- temp$coefficients[2, 2]
        results$t_value[x] <- temp$coefficients[2, 3]
        results$p_values[x] <- temp$coefficients[2, 4]
        results$r_sq[x] <- temp$adj.r.squared
      }
      
      ## multiple testing correlation by fdr? 
      results <- as.data.frame(results)
      results$FDR <- p.adjust(results$p_values, method = "fdr")
      results <- results[order(results$FDR, decreasing = FALSE), ]
      
      #print(paste0("finished analysing column ", y))
      immuno[[col_names[y]]] <- results
    }
    
    ## random test, let's use the first row of the dataframe against tls 
    check_test <- lm(immuno_data[, 3]~cnv[1, ]+pheno$age+pheno$stage, data = as.data.frame(cnv))
    check_test <- summary(check_test)
  }
  
  ## check test 
  print(paste0("this is a check test using first row of the cnv data and tls for ", cancer))
  print(check_test)
  
  return(immuno)
}
```


#### Application      
```{r application_func, warning=FALSE}
## this is a function to apply the spearman or linear regression function 
apply_func <- function(cancer, focal_name, broad_name, test_name){
  ## focal_name = "result_imm_cor_f_" or "result_imm_lm_f_"
  ## broad_name = "result_imm_cor_b_" or "result_imm_lm_b_"
  ## test = either linear_regression or spearman 
  
  result_immune_f <- statistics_test(cancer, focal = TRUE, test = test_name)
  result_immune_b <- statistics_test(cancer, broad = TRUE, test = test_name)
  
  name_f <- make.names(paste0(focal_name, cancer))
  name_b <- make.names(paste0(broad_name, cancer))
  
  assign(name_f, result_immune_f, envir = .GlobalEnv)
  assign(name_b, result_immune_b, envir = .GlobalEnv)
}


## this is a function to print the resulting dataframe 
print_df <- function(cancer, focal_name, broad_name){
  ## broad_name = "result_imm_cor_b_" or "result_imm_lm_b_"
  ## focal_name = "result_imm_cor_f_" or "result_imm_lm_f_"
  
  broad_df <- get(paste0(broad_name, cancer))
  print(paste0("the immuno-association analysis of ", cancer, " broad_gistic2"))
  names_list <- names(broad_df)
  invisible(lapply(seq_along(broad_df), function(cancer) {
    cat(paste("\n---", names_list[cancer], "---\n"))
    print(head(broad_df[[cancer]], 3))
  }))
  
  focal_df <- get(paste0(focal_name, cancer))
  print(paste0("the immuno-association analysis of ", cancer, " focal_gistic2"))
  names_list <- names(focal_df)
  invisible(lapply(seq_along(focal_df), function(cancer) {
    cat(paste("\n---", names_list[cancer], "---\n"))
    print(head(focal_df[[cancer]], 3))
  }))
}
```


```{r run, warning=FALSE}
for(i in TCGA){
  apply_func(cancer = i, focal_name = "result_imm_cor_f_", broad_name = "result_imm_cor_b_", test_name = "spearman")
  apply_func(cancer = i, focal_name = "result_imm_lm_f_", broad_name = "result_imm_lm_b_", test_name = "linear_regression")
}

#for(i in TCGA){
  #print_df(cancer = i, focal_name = "result_imm_cor_f_", broad_name = "result_imm_cor_b_")
  #print_df(cancer = i, focal_name = "result_imm_lm_f_", broad_name = "result_imm_lm_b_")
#}
```

### Application of SCNA scores and Aneuploidy score against immune phenotype      
#### Read in and combine SCNA acores with Aneuploidy Scores       
```{r read_SCNA_scores}
## function: resolve_duplication
resolve_dup_name <- function(file, colname){
  file <- file %>% group_by_at(colname) %>% summarise_all(mean, na.rm = TRUE) %>% ungroup()
  file <- as.data.frame(file)
  
  return(file)
}

## read in files 
for(i in TCGA){
  file <- read_csv(paste0(SCNA, "/tcga/", i, "/normalised_SCNA_score.csv"), show_col_types = FALSE)
  file_name <- make.names(paste0(i, "_SCNA_score"))
  file <- as.data.frame(file)
  file <- resolve_dup_name(file, "...1")
  rownames(file) <- file$...1
  file$...1 <- NULL
  
  assign(file_name, file, envir = .GlobalEnv)
}
```
#### Read in Aneuploidy Score 
```{r aneuploidy_score}
URL <- "https://ars.els-cdn.com/content/image/1-s2.0-S1535610818301119-mmc2.xlsx"
aneuploidy_score <- paste0(GISTIC2, "/aneuploidy_score.xlsx")

if(!file.exists(aneuploidy_score)){
  download.file(URL, destfile = paste0(GISTIC2, "/aneuploidy_score.xlsx"), method = "libcurl")
} 

aneuploidy_score <- read_excel(paste0(GISTIC2, "/aneuploidy_score.xlsx"), skip = 1, col_names = TRUE)

for(i in TCGA){
  sample_name <- get(i, envir = .GlobalEnv)
  name <- make.names(paste0(i, "_aneuploidy_score"))
  score <- aneuploidy_score %>% filter(Sample %in% sample_name)
  score <- score[, c(1, 3:ncol(score))]
  score <- resolve_dup_name(score, "Sample")

  assign(name, score)
}
```

#### Combine 
```{r}
scna_as_combine <- function(cancer, columns){
  ## columns = c("rank_norm_focal", "rank_norm_chrom", "rank_norm_arm", "sum_score", "AneuploidyScore(AS)", "Leuk", "Purity")
  
  scna <- get(paste0(cancer, "_SCNA_score"), envir = .GlobalEnv)
  as <- get(paste0(cancer, "_aneuploidy_score"), envir = .GlobalEnv)
  
  rownames(as) <- as$Sample
  as$Sample <- NULL
  
  scna_as <- merge(scna, as, by = "row.names", all.x = TRUE)
  rownames(scna_as) <- scna_as$Row.names
  scna_as$Row.names <- NULL
  
  scna_as <- scna_as[, columns]
  
  scna_as_name <- make.names(paste0(cancer, "_scna_as"))
  assign(scna_as_name, scna_as, envir = .GlobalEnv)
}

keep = c("rank_norm_focal", "rank_norm_chrom", "rank_norm_arm", "sum_score", "AneuploidyScore(AS)", "Leuk", "Purity")
for(i in TCGA){
  scna_as_combine(i, columns = keep)
}
```

```{r warning=FALSE}
for(i in TCGA){
  result_scna_lm_lung <- statistics_test(i, not_gistic = TRUE, test = "linear_regression")
  result_scna_lm_ova <- statistics_test(i, not_gistic = TRUE, test = "linear_regression")
  result_scna_lm_mel <- statistics_test(i, not_gistic = TRUE, test = "linear_regression")
}
```


#### Others 
```{r save_image}
save.image(file = "/rds/general/user/ph323/home/MRes_project_1/Codes/2_immune_association/immune_association.RData")
```


